The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-04-22 paper.qmd
edits to the grammar of graphics section

Unlike uncertainty, the concept of the "information" in a plot is not entirely avoided and works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot. When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005]. First the mathematical information in a plot is established in the "varset" steps, and then that mathematical information is visually represented, which is defined in the "graph" steps. Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymmetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical. However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage. Additionally, the grammar of graphics was not designed to extend to "uncertainty visualisations" as @Leland2005 referred to uncertainty as "semantics", so concepts that are common to uncertainty mathematically, such as resampling do not quite fit within the framework. For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line. These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used). This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework. Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable. @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcomes plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.

@@

>>>>

Works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot, and in doing so allow plots to be considered a statistic [@Vanderplas2020]. This statistic that *The Grammar of Graphics* defines fundamentally a descriptive statistic. @fig-pipeline, and contains the set of necessary and sufficient steps that are required to make a graphic, as defined in *The Grammar of Graphics* [@Leland2005]. It can be seen that the grammar assumes we have our data, performs a set of functions on that data, and produces a statistic. The grammar of graphics fundamentally describes the graphic as it is, not as it *would* be were our sample size to increase to infinity. As a matter of fact, the graphics assumption that we start from a data set that has an unchanging sample size is at odds with inference that typically involves assuming an infinitely large sample size. @fig-infintegraph shows that as n approaches infinity, two graphics that are very similar according to the *The Grammar of Graphics* can diverge into different graphics, while two graphics that are very different, can converge into the same graphic. *The Grammar of Graphics* itself does not recognize this issue, as uncertainty is referred to as "semantics" and is ultimately hand waved away. This does not mean describing a graphic using inferential statistics is completely unheard of. Some graphical descriptors, such as Product Plots [@Wickham2011] describe graphics using the underlying distribution, however the framework is not a substitute for *The Grammar of Graphics*. Other authors have noticed difficulties with inferential statistics and and created an addition to the grammar that should make depicting a distribution easier [@Pu2020; @Kay2023]. Unfortunately, this addition falls short as it does not recognize *why* latent distributions are difficult to describe using *The Grammar of Graphics* and carries over many of the same problems.

This does not mean *The Grammar of Graphics* should be abandoned when we visualize uncertainty, rather, these difficulties are something we should keep in mind when

If uncertainty is fundamentally a by-product of inference, then the visualisations as described by *The Grammar of Graphics*

Most uncertainty visualisation experiments are written with a seeming indifference to the grammar of Graphics, despite it being necessary to untangle experimental results. The primary reason for this seems to be the absent definition of uncertainty. Uncertainty is fundamentally related to inference and descriptive statistics cannot have uncertainty.

Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymmetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical. However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage. , such as resampling do not quite fit within the framework. For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line. These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used). This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework. Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable. @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcomes plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.

@@

 ![(infintiegraph](infinitegraph.png){#fig-infinitegraph}
